Semistochastic importance sampling of second-quantized operators in determinant space

### 2021-06-07 notes/talk/talk.tex
Final version of talk

\section{Motivation}
\begin{frame}{Types of quantum chemistry algorithms}
	\begin{enumerate}
		\item Orbital-space algorithms:
		\begin{enumerate}
			\item Coupled cluster
			\item Moller-Plesset perturbation theory
			\linebreak
		\end{enumerate}
		\item Determinant-space algorithms:
		\begin{enumerate}
			\item Full CI
			\item Selected CI
			\item Epstein-Nesbet perturbation theory
			\item Full CI Quantum Monte Carlo (FCIQMC)
		\end{enumerate}
	\end{enumerate}
\end{frame}

@@

\begin{frame}{Approaches for computing $H\left|D_i\right\rangle$}
	\small
	Broadly, there are three main approaches:
	\begin{enumerate}
		\item {\bf Deterministic:} Simply evaluating all single and double excitations from the initial Slater determinant. Complexity: $\mathcal{O}(N^2 (M-N)^2)$
		\item {\bf Stochastic:} Sampling excitations according to some distribution. Reduces the complexity, but introduces a stochastic uncertainty (and in projector methods, the fermion sign problem!)
		\item {\bf Semistochastic:} Evaluating the largest-magnitude components, and sampling the remaining, smaller components. Reduces complexity relative to the deterministic approach, but with a greatly reduced stochastic uncertainty relative to fully stochastic methods. In projector methods, it also mitigates the bias incurred in taming the fermion sign problem, such as the initiator bias in FCIQMC.
	\end{enumerate}
\end{frame}

@@

\section{Algorithm}
\subsection{Deterministic component}
\begin{frame}{Deterministic component}
	\Wider{
		\begin{itemize}
			\item  \textbf{Setup:}
			\begin{itemize}
				\item  For each pair of orbitals $\left\{p,q\right\}$, store a list of %triplets
				$\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$,
				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
				\linebreak
				\item Also compute (for sampling algorithm):
				\begin{eqnarray}
				\sum_{rs\in\textrm{remaining}} |H(rs\leftarrow pq)|
				\end{eqnarray}
			\end{itemize}
			{\color{white}  skip line}
			\\
			\uncover<2->{
				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|>\epsilon$): %for at least one $i\in \mathcal{V}$)
				\\{\color{white}  skip line}
				\\
			}
			\begin{itemize}
				\uncover<3->{
					%\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
					%\\
					\item For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
					of triplets $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
					through the list until a triplet is reached for which $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
				}
				\\{\color{white}  skip line}
				\\
			\end{itemize}
		\end{itemize}
	}
\end{frame}

@@

\subsection{Details and extensions}
\begin{frame}{What about single excitations?}
	\begin{itemize}
		%\small
		\item Unlike double excitations, single excitation matrix element magnitudes depend on the orbital occupancies of the exciting determinant
		\item But we can compute upper bounds to the single excitation magnitudes:
%		\tiny
		\begin{eqnarray}
			\left|H(p\rightarrow r)\right| %&=& \left|f_{pr} + \sum_{q\in \textrm{occ}} g_{pqqr}\right|\\
			\le \left|H(p\rightarrow r)\right|_{\max},%&=&\max\Bigg( \left|f_{pr} + \sum_{q\in [N-1 \textrm{ largest}]} g_{pqqr}\right|,
%			\\&&\quad \quad \quad 
%			\quad \left|f_{pr} + \sum_{q\in [N-1 \textrm{ smallest}]} g_{pqqr}\right|\Bigg),
		\end{eqnarray}
%		%\]
%		\small
%		where the sums in the last line above are over the $N-1$ distinct spin-orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively
		where $|H(p\rightarrow r)|_{\max}$ is the largest magnitude the $p\rightarrow r$ excitation can have from any initial determinant
		\item The simplest way to incorporate singles is to sample them with probability $P(|H(p\rightarrow r)|) \propto f(|H(p\rightarrow r)|_{\max})$ and skip them in the deterministic step
	\end{itemize}
\end{frame}

\begin{frame}{Sampling `invalid' excitations}
\begin{itemize}
	\item Sometimes this algorithm will sample `invalid' excitations to already-occupied orbitals
	\item While unbiased, this increases the variance in the samples
	\item We can fix this as follows:
	\begin{enumerate}
		\item If an `invalid' excitation is sampled, discard it and resample until a valid excitation is sampled
		\item Scale the sample probabilities by a constant factor so they remain normalized (can be performed in $\mathcal{O}(N^3)$ time during deterministic step)
	\end{enumerate}
\end{itemize}	
\end{frame}

@@

\begin{frame}{Semistochastic Epstein-Nesbet Perturbation Theory}
	\begin{enumerate}
		\item Divide $V\left|\psi_0\right\rangle$ into two pieces:
		\begin{eqnarray}
			H\left|\psi_0\right\rangle = \left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon} + \left(V\left|\psi_0\right\rangle\right)_{<\epsilon}.
		\end{eqnarray}
		\item Estimate $\Delta E$ deterministically:
		\begin{eqnarray}
			\Delta E\left[V\left|\psi_0\right\rangle\right] \approx \Delta E\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right] 
		\end{eqnarray}
		\item Stochastically evaluate the difference between the exact and approximate $\Delta E$
		%\begin{eqnarray}
		%	\Delta E\left[V\left|\psi_0\right\rangle\right] - \Delta E\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right] .
		%\end{eqnarray}
	\end{enumerate}
\end{frame}

@@

\begin{frame}{Sampling algorithm in Epstein-Nesbet PT}
	\begin{itemize}
		\item New approach (Importance sampling):
		\linebreak
		\begin{itemize}
			\item Sample pairs of determinants $\{D_i, D_a\}$ %from $\psi_0$ and $\left(H\left|\psi_0\right\rangle\right)_{<\epsilon}$ respectively
			using importance sampling: \begin{eqnarray} P(D_i, D_a) \propto \left(H_{ai} c_i\right)^2 \end{eqnarray}
			\item Use those samples to evaluate the stochastic component of $\Delta E$,%\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right]$,
			whose largest-magnitude component is
			\begin{eqnarray}
				\sum_{ai %\substack{ai\\\textrm{ s.t. }\\|H_{ai}c_i|<\epsilon}
				}\frac{ \left(H_{ai}c_i\right)^2}{E_0 - E_a}
			\end{eqnarray}
			\item Advantages:
			\begin{itemize}
				\item Uses importance sampling to greatly reduce the variance!
				\item No time or memory is wasted on deterministically evaluating contributions from small-magnitude $H_{ai}$ elements!
			\end{itemize}
		\end{itemize}
	\end{itemize}
\end{frame}

>>>>

\section{Motivation}
\begin{frame}{Types of quantum chemistry algorithms}
	\begin{enumerate}
		\item Operator-based algorithms:
		\begin{enumerate}
			\item Coupled cluster
			\item Moller-Plesset perturbation theory
			\linebreak
		\end{enumerate}
		\item Wavefunction-based/determinant-space algorithms:
		\begin{enumerate}
			\item Full CI
			\item Selected CI
			\item Epstein-Nesbet perturbation theory
			\item Full CI Quantum Monte Carlo (FCIQMC)
		\end{enumerate}
	\end{enumerate}
\end{frame}

@@

\begin{frame}{Approaches for computing $H\left|D_i\right\rangle$}
	\small
	Broadly, there are three main approaches:
	\begin{enumerate}
		\item {\bf Deterministic:} Simply evaluating all single and double excitations from the initial Slater determinant. Complexity: $\mathcal{O}(N^2 (M-N)^2)$
		\item {\bf Stochastic:} Sampling excitations according to some distribution. Reduces the complexity, but introduces a stochastic uncertainty (and in projector methods, the fermion sign problem!)
		\item {\bf Semistochastic:} Evaluating the largest-magnitude components, and sampling the remaining, smaller components. Reduces complexity relative to the deterministic approach, but with a \textbf{greatly reduced stochastic uncertainty} relative to fully stochastic methods. In projector methods, it also mitigates the bias incurred in taming the fermion sign problem, such as the initiator bias in FCIQMC.
	\end{enumerate}
\end{frame}

@@

\section{Algorithm}
\subsection{Deterministic component}
\begin{frame}{Deterministic component}
	\Wider{
		\begin{itemize}
			\item  \textbf{Setup:}
			\begin{itemize}
				\item  For each pair of orbitals $\left\{p,q\right\}$, store a list of %triplets
				$\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$,
				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
				\linebreak
				\item Also compute (for sampling algorithm):
				\begin{eqnarray}
				\sum_{rs\in\textrm{remaining}} |H(rs\leftarrow pq)|
				\end{eqnarray}
			\end{itemize}
			{\color{white}  skip line}
			\\
			\uncover<2->{
				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|\ge\epsilon$): %for at least one $i\in \mathcal{V}$)
				\\{\color{white}  skip line}
				\\
			}
			\begin{itemize}
				\uncover<3->{
					%\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
					%\\
					\item For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
					of $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
					until $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
				}
				\\{\color{white}  skip line}
				\\
			\end{itemize}
		\end{itemize}
	}
\end{frame}

@@

\subsection{Details and extensions}
\begin{frame}{What about single excitations?}
	\begin{itemize}
		%\small
		\item Unlike double excitations, single excitation matrix element magnitudes depend on the orbital occupancies of the exciting determinant
		\item But we can compute upper bounds to the single excitation magnitudes:
%		\tiny
		\begin{eqnarray}
			\left|H(p\rightarrow r)\right| %&=& \left|f_{pr} + \sum_{q\in \textrm{occ}} g_{pqqr}\right|\\
			\le \left|H(p\rightarrow r)\right|_{\max},%&=&\max\Bigg( \left|f_{pr} + \sum_{q\in [N-1 \textrm{ largest}]} g_{pqqr}\right|,
%			\\&&\quad \quad \quad 
%			\quad \left|f_{pr} + \sum_{q\in [N-1 \textrm{ smallest}]} g_{pqqr}\right|\Bigg),
		\end{eqnarray}
%		%\]
%		\small
%		where the sums in the last line above are over the $N-1$ distinct spin-orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively
		where $|H(p\rightarrow r)|_{\max}$ is the largest magnitude the $p\rightarrow r$ excitation can have from \textit{any} initial determinant
		\item The simplest way to incorporate singles is to sample them with probability $P(|H(p\rightarrow r)|) \propto f(|H(p\rightarrow r)|_{\max})$ and skip them in the deterministic step
	\end{itemize}
\end{frame}

\begin{frame}{Sampling `invalid' excitations}
\begin{itemize}
	\item Sometimes this algorithm will sample `invalid' excitations to already-occupied orbitals
	\item While unbiased, this increases the variance in the samples
	\item We can fix this as follows:
	\begin{enumerate}
		\item If an `invalid' excitation is sampled, discard it and resample until a valid excitation is sampled
		\item Scale the sample probabilities by a constant factor so they remain normalized (can be performed in $\mathcal{O}(N^4 + N_{\textrm{det}} N^3)$ time during deterministic step)
	\end{enumerate}
\end{itemize}	
\end{frame}

@@

\begin{frame}{Semistochastic Epstein-Nesbet Perturbation Theory}
	\begin{enumerate}
		\item Divide $V\left|\psi_0\right\rangle$ into two pieces:
		\begin{eqnarray}
			V\left|\psi_0\right\rangle = \left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon} + \left(V\left|\psi_0\right\rangle\right)_{<\epsilon}.
		\end{eqnarray}
		\item Estimate $\Delta E$ deterministically:
		\begin{eqnarray}
			\Delta E\left[V\left|\psi_0\right\rangle\right] \approx \Delta E\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right] 
		\end{eqnarray}
		\item Stochastically evaluate the difference between the exact and approximate $\Delta E$
		%\begin{eqnarray}
		%	\Delta E\left[V\left|\psi_0\right\rangle\right] - \Delta E\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right] .
		%\end{eqnarray}
	\end{enumerate}
\end{frame}

@@

\begin{frame}{Sampling algorithm in Epstein-Nesbet PT}
	\begin{itemize}
		\item New approach (Importance sampling):
		\linebreak
		\begin{itemize}
			\item Sample pairs of determinants $\{D_i, D_a\}$ %from $\psi_0$ and $\left(H\left|\psi_0\right\rangle\right)_{<\epsilon}$ respectively
			using importance sampling: \begin{eqnarray} P(D_i, D_a) \propto \left(H_{ai} c_i\right)^2 \end{eqnarray}
			\item Use those samples to evaluate the stochastic component of $\Delta E$, %\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right]$,
			whose largest-magnitude component is
			\begin{eqnarray}
				\sum_{ai %\substack{ai\\\textrm{ s.t. }\\|H_{ai}c_i|<\epsilon}
				}\frac{ \left(H_{ai}c_i\right)^2}{E_0 - E_a}
			\end{eqnarray}
			\item Advantages:
			\begin{itemize}
				\item Uses importance sampling to greatly reduce the variance!
				\item No time or memory is wasted on deterministically evaluating contributions from small-magnitude $H_{ai}$ elements!
			\end{itemize}
		\end{itemize}
	\end{itemize}
\end{frame}
